cardinalR: Generating interesting high-dimensional data structures

### 2025-09-23 paper-cardinalR.Rmd
Updated this with new org functions
Updated Di's comment on abstract first sentence

---
title: "cardinalR: Generating interesting high-dimensional data structures"
abstract: >
  A high-dimensional dataset is where each observation is described by many features, or dimensions. Such a dataset might contain various types of structures that have complex geometric properties, such as nonlinear manifolds, clusters, or sparse distributions. We can generate data containing a variety of structures using mathematical functions and statistical distributions. Sampling from a multivariate normal distribution will generate data in an elliptical shape. Using a trigonometric function we can generate a spiral. A torus function can create a donut shape. High-dimensional data structures are useful for testing, validating, and improving algorithms used in dimensionality reduction, clustering, machine learning, and visualization. Their controlled complexity allows researchers to understand challenges posed in data analysis and helps to develop robust analytical methods across diverse scientific fields like bioinformatics, machine learning, and forensic science. Functions to generate a large variety of structures in high dimensions are organized into the R package `cardinalR`, along with some already generated examples.
draft: true
author:  
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
output: 
 rjtools::rjournal_article:
    keep_md: true
bibliography: paper-cardinalR.bib
editor_options: 
  chunk_output_type: inline
---

@@

High-dimensional generalizations are provided by `gen_orglinearbranches(n, p, k)` and `gen_orgcurvybranches(n, p, k)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) d-e"} else { "Figure \\@ref(fig:branch-proj) d-e"})`). Each branch is embedded in a unique or repeated $2\text{-}D$ subspace of the $p\text{-}D$ space. Linear branches follow

>>>>

---
title: "cardinalR: Generating interesting high-dimensional data structures"
abstract: >
  A high-dimensional dataset is one where each observation is described by many features, or dimensions, with associations between them. Such a dataset might contain various types of structures that have complex geometric properties, such as nonlinear manifolds, clusters, or sparse distributions. We can generate data containing a variety of structures using mathematical functions and statistical distributions. Sampling from a multivariate normal distribution will generate data in an elliptical shape. Using a trigonometric function we can generate a spiral. A torus function can create a donut shape. High-dimensional data structures are useful for testing, validating, and improving algorithms used in dimensionality reduction, clustering, machine learning, and visualization. Their controlled complexity allows researchers to understand challenges posed in data analysis and helps to develop robust analytical methods across diverse scientific fields like bioinformatics, machine learning, and forensic science. Functions to generate a large variety of structures in high dimensions are organized into the R package `cardinalR`, along with some already generated examples.
draft: true
author:  
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
output: 
 rjtools::rjournal_article:
    keep_md: true
bibliography: paper-cardinalR.bib
editor_options: 
  chunk_output_type: inline
---

@@

High-dimensional generalizations are provided by `gen_orglinearbranches(n, p, k)` and `gen_orgcurvybranches(n, p, k)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) d-e"} else { "Figure \\@ref(fig:branch-proj) d-e"})`). Each branch is embedded in a unique or repeated $2\text{-}D$ subspace of the $p\text{-}D$ space. When `allow_share = TRUE`, multiple branches may share the same subspace; otherwise, subspaces are sampled without replacement until all possible $\binom{p}{2}$ combinations are exhausted, after which additional branches may repeat subspaces. Linear branches follow

### 2025-09-23 paper-cardinalR.tex
Updated this with new org functions
Updated Di's comment on abstract first sentence

\abstract{%
A high-dimensional dataset is where each observation is described by many features, or dimensions. Such a dataset might contain various types of structures that have complex geometric properties, such as nonlinear manifolds, clusters, or sparse distributions. We can generate data containing a variety of structures using mathematical functions and statistical distributions. Sampling from a multivariate normal distribution will generate data in an elliptical shape. Using a trigonometric function we can generate a spiral. A torus function can create a donut shape. High-dimensional data structures are useful for testing, validating, and improving algorithms used in dimensionality reduction, clustering, machine learning, and visualization. Their controlled complexity allows researchers to understand challenges posed in data analysis and helps to develop robust analytical methods across diverse scientific fields like bioinformatics, machine learning, and forensic science. Functions to generate a large variety of structures in high dimensions are organized into the R package \texttt{cardinalR}, along with some already generated examples.
}

@@

High-dimensional generalizations are provided by \texttt{gen\_orglinearbranches(n,\ p,\ k)} and \texttt{gen\_orgcurvybranches(n,\ p,\ k)} (Figure \ref{fig:branch-proj} d-e). Each branch is embedded in a unique or repeated \(2\text{-}D\) subspace of the \(p\text{-}D\) space. Linear branches follow

>>>>

\abstract{%
A high-dimensional dataset is one where each observation is described by many features, or dimensions, with associations between them. Such a dataset might contain various types of structures that have complex geometric properties, such as nonlinear manifolds, clusters, or sparse distributions. We can generate data containing a variety of structures using mathematical functions and statistical distributions. Sampling from a multivariate normal distribution will generate data in an elliptical shape. Using a trigonometric function we can generate a spiral. A torus function can create a donut shape. High-dimensional data structures are useful for testing, validating, and improving algorithms used in dimensionality reduction, clustering, machine learning, and visualization. Their controlled complexity allows researchers to understand challenges posed in data analysis and helps to develop robust analytical methods across diverse scientific fields like bioinformatics, machine learning, and forensic science. Functions to generate a large variety of structures in high dimensions are organized into the R package \texttt{cardinalR}, along with some already generated examples.
}

@@

High-dimensional generalizations are provided by \texttt{gen\_orglinearbranches(n,\ p,\ k)} and \texttt{gen\_orgcurvybranches(n,\ p,\ k)} (Figure \ref{fig:branch-proj} d-e). Each branch is embedded in a unique or repeated \(2\text{-}D\) subspace of the \(p\text{-}D\) space. When \texttt{allow\_share\ =\ TRUE}, multiple branches may share the same subspace; otherwise, subspaces are sampled without replacement until all possible \(\binom{p}{2}\) combinations are exhausted, after which additional branches may repeat subspaces. Linear branches follow
